03-TTS-travel-to-school-impedance

### 2024-04-09 data-prep/03-travel-times-TTS.qmd
Draft of 'Manuscript.qmd' for special issue

## Preliminaries

Load packages:
```{r load-packages}
library(readxl)
library(sf)
library(tidyverse)
library(units)
library(tmap)
library(ggplot2)
```

@@

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp")
```

Project object:
```{r}
ggh_taz <- ggh_taz %>%
  st_transform(crs = 32617)
```

Prepare traffic analysis zones:
```{r}
ggh_taz <- ggh_taz %>%
  transmute(GTA06 = as.character(GTA06),
            AREA = st_area(geometry) %>% 
              set_units(km^2) %>% 
              drop_units())
```

Extract the IDs for all TAZ:
```{r}
ghta_taz_id <- ggh_taz$GTA06
```

@@

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_pd <- st_read("data-inputs/TTS/tts06_pd_83.shp")
```

Project object:
```{r}
ggh_pd <- ggh_pd %>%
  st_transform(crs = 32617)
```

@@

od_trips_2016_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)
# od_trips_2011 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2011.txt", 
#                       delim = "\t", 
#                       col_names = FALSE)
# od_trips_2016 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2016.txt", 
#                       delim = "\t", 
#                       col_names = FALSE)

Find the position in the table where the information on work trips begins:
```{r}
# idx <- which(od_trips_2011$X1 == "TABLE    : stu_stat (Full time)")
idx <- which(od_trips_2011_motor$X1 == "COLUMN : gta06_sch")
```

@@

Now let's see how many students travel outside of their catchment. 
```{r}
#load catchments
load(file="data-products/ALL_CATCHS.RData")
CATCH_2011_Public <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Public") %>% st_set_crs(st_crs(ham_taz))
CATCH_2011_Cath <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Public <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Public")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Cath <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))

@@

```{r}
## THIS FUNCTION WAS CREATED USING CHATGDPT on DEC 6 2023

# Function to check if values in a row of df1 are within any list in df2
check_within_lists <- function(df1, df2) {
  # Initialize an empty list to store the results
  result_list <- list()

@@

Libraries:
```{r setup, include=FALSE, eval=FALSE}
library(disk.frame)
library(ggplot2)
library(gridExtra)
library(httr)
library(knitr)
#install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was receiving errors
library(kableExtra)
#install.packages("devtools")
#install.packages("Rtools")
#devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
library(lodes)
library(progress)
library(purrr)
library(r5r)
#library(stplanr)
library(tidyr)
library(tmap)
library(zoo) # for rollapplyr

# setup for disk.frame
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx512m")
options(future.globals.maxSize = Inf)
```

Format the destination and origin points (ID, Long, Lat):
```{r, eval=FALSE}
ham_taz_coords <-  
  cbind(ham_taz %>% st_drop_geometry()%>% transmute(GTA06), st_centroid(ham_taz) %>% st_transform(crs = 4326) %>% st_coordinates()) %>%
  rename(lon = "X", lat = "Y", id = "GTA06") 
```
Set-up directories and download the osm network:
```{r set up r5 path, include=FALSE, eval=FALSE}
# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
dir.create("data-inputs/TTS/r5_graph")
r5_path <- file.path("data-inputs/TTS/r5_graph")
```

@@

```
Set Up R5 Routing:
```{r build graph, include = FALSE, eval=FALSE}
# NOTE: I had signficant issues getting this function to run. Since the most recent r5r update (before Nov ~25 2022 when I was working on this), I believe the setup_r5 requires smaller road networks. So putting the full ontario geofabrik extract no longer works for me. I used http://share.mapbbcode.org/ to export the Hamilton area, and then OSMCONVERT to convert the .osm file into a .pbf.
r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE, overwrite = TRUE, temp_dir = TRUE)

@@

# create origin-destination pairs
origins_chunks <- as.disk.frame(ham_taz_coords,
                          outdir = "data-inputs/df/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

@@

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car <- as.data.frame(output_df)

save("output_OD_car", file = "data-inputs/TTS/output_OD_car.Rdata")
#I only made the cut off 60 minutes, so if some parcels do not have a travel time, rerun this for a 120 min cut off. The following are 120mins cutoff
```

@@

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk <- as.data.frame(output_df)

save("output_OD_walk", file = "data-inputs/TTS/output_OD_walk.Rdata")
```
Java garbage collector:
```{r, eval=FALSE}
stop_r5(r5_HAM)
rJava::.jgc(R.gc = TRUE)
```

## Inspecting the travel times and calculating TLD
```{r}
load(file = "data-inputs/TTS/output_OD_car.Rdata")
summary(output_OD_car)
#we'll set 0 travel times to equal 7 min for inter-zonal travel, half of the 1st Qu. travel time.

load(file = "data-inputs/TTS/output_OD_walk.Rdata")
summary(output_OD_walk) #max trip length calculated is 27 mins -- this corresponds to 'within catchment'. lets make 0 min trips into half of 1st Q. so 4min.
```

Trip length distribution (cutting the travel_time into 150 intervals and summing all trips within each interval). The TLD is a probability density function for likelihood of travel informed by travel time:
```{r}
#join travel_time to 2011 and 2016 OD trips
od_trips_2011_motor <- od_trips_2011_motor %>% 
  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

od_trips_2016_motor <- od_trips_2016_motor %>% 
  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id"))  %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

tld_2011_nonmotor <- od_trips_2011_nonmotor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2016_nonmotor <- od_trips_2016_nonmotor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))
```

@@

all_tt_2011_nonmotor <- od_trips_2011_nonmotor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2011_nonmotor$Trips))
all_tt_2011_nonmotor <- all_tt_2011_nonmotor[rep(seq_len(dim(all_tt_2011_nonmotor)[1]), all_tt_2011_nonmotor$Trips), 2]

@@

all_tt_2016_nonmotor <- od_trips_2016_nonmotor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2016_nonmotor$Trips))
all_tt_2016_nonmotor <- all_tt_2016_nonmotor[rep(seq_len(dim(all_tt_2016_nonmotor)[1]), all_tt_2016_nonmotor$Trips), 2]

@@

And save the gamma functions, to use in the analysis
```{r}
save(gamma2011_motor, file="data-inputs/TTS/gamma2011_motor.Rdata")
save(gamma2016_motor, file="data-inputs/TTS/gamma2016_motor.Rdata")
save(exp2011_nonmotor, file="data-inputs/TTS/exp2011_nonmotor.Rdata")
save(exp2016_nonmotor, file="data-inputs/TTS/exp2016_nonmotor.Rdata")
```

>>>>

---
title: "03-TTS-travel-to-school-impedance"
format: html
---
Load packages:
```{r load-packages}
library(readxl)
library(sf)
library(tidyverse)
library(units)
library(tmap)
library(ggplot2)
```

@@

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") %>%   st_transform(crs = 32617)
```

Prepare traffic analysis zones:
```{r}
ggh_taz <- ggh_taz %>%
  transmute(GTA06 = as.character(GTA06),
            AREA = st_area(geometry) %>% 
              set_units(km^2) %>% 
              drop_units())
#Extract the IDs for all TAZ:
ghta_taz_id <- ggh_taz$GTA06
```

@@

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_pd <- st_read("data-inputs/TTS/tts06_pd_83.shp") %>%
  st_transform(crs = 32617)
```

@@

od_trips_2016_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)
```

Find the position in the table where the information on work trips begins:
```{r}
idx <- which(od_trips_2011_motor$X1 == "COLUMN : gta06_sch")
```

@@

Now let's see how many students travel outside of their catchment. 
```{r} 
#load catchments - available through the hamONdests package (https://github.com/soukhova/hamONdests):
library(hamONdests)
data(School_Catchments_2011_2016)

CATCH_2011_Public <- School_Catchments_2011_2016%>%filter(Year == 2011 & Level == "Elementary" & System == "Public") %>% st_set_crs(st_crs(ham_taz))
CATCH_2011_Cath <- School_Catchments_2011_2016%>%filter(Year == 2011 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Public <- School_Catchments_2011_2016%>%filter(Year == 2016 & Level == "Elementary" & System == "Public")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Cath <- School_Catchments_2011_2016%>%filter(Year == 2016 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))

@@

```{r}
## THIS FUNCTION WAS CREATED USING CHATGDPT on DEC 6 2023 -- this is a function to check if values in a row of df1 are within any list in df2
check_within_lists <- function(df1, df2) {
  # Initialize an empty list to store the results
  result_list <- list()

@@

Libraries:
```{r setup, include=FALSE, eval=FALSE}
library(disk.frame)
library(tidyverse)
library(r5r)
library(sf)
library(tmap)

# setup for disk.frame
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx6G")
```

Format the destination and origin points (ID, Long, Lat):
```{r, eval=FALSE}
ham_taz_coords <-  
  cbind(ham_taz %>% st_drop_geometry()%>% transmute(GTA06), st_centroid(ham_taz) %>% st_transform(crs = 4326) %>% st_coordinates()) %>%
  rename(lon = "X", lat = "Y", id = "GTA06") 
```
Set-up directories and download the osm network:
```{r set up r5 path, include=FALSE, eval=FALSE}
# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
dir.create("data-inputs/Travel-times")
r5_path <- file.path("data-inputs/Travel-times")
```

@@

```
Set Up R5 Routing:
```{r build graph, include = FALSE, eval=FALSE}
# NOTE: I had signficant issues getting this function to run. Since the most recent r5r update (before Nov ~25 2022 when I was working on this), I believe the setup_r5 requires smaller road networks. So putting the full ontario geofabrik extract no longer works for me. I used http://share.mapbbcode.org/ to export the Hamilton area, and then OSMCONVERT to convert the .osm file into a .pbf.
r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE, overwrite = TRUE, temp_dir = TRUE)
```

@@

# create origin-destination pairs
origins_chunks <- as.disk.frame(ham_taz_coords,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

@@

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car <- as.data.frame(output_df)
save("output_OD_car", file = "data-inputs/TTS/TTS_tt_car.Rdata")
#I only made the cut off 60 minutes, so if some parcels do not have a travel time, rerun this for a 120 min cut off. The following are 120mins cutoff
```

@@

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk <- as.data.frame(output_df)
save("output_OD_walk", file = "data-inputs/TTS/TTS_tt_walk.Rdata")
```
Java garbage collector:
```{r, eval=FALSE}
stop_r5(r5_HAM)
rJava::.jgc(R.gc = TRUE)
```

## Inspecting the travel times and calculating TLD
```{r}
load(file = "data-inputs/TTS/TTS_tt_car.Rdata")
TTS_tt_car <- output_OD_car
rm("output_OD_car")
summary(TTS_tt_car)
#we'll set 0 travel times to equal 7 min for inter-zonal travel, half of the 1st Qu. travel time.

load(file = "data-inputs/TTS/TTS_tt_walk.Rdata")
TTS_tt_walk <- output_OD_walk
rm("output_OD_walk")
summary(TTS_tt_walk) #max trip length calculated is 27 mins -- this corresponds to 'within catchment'. lets make 0 min trips into half of 1st Q. so 4min.
```

Trip length distribution (cutting the travel_time into 150 intervals and summing all trips within each interval). The TLD is a probability density function for likelihood of travel informed by travel time:
```{r}
#join travel_time to 2011 and 2016 OD trips
od_trips_2011_motor <- od_trips_2011_motor %>% 
  merge(TTS_tt_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

od_trips_2016_motor <- od_trips_2016_motor %>% 
  merge(TTS_tt_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
  merge(TTS_tt_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  merge(TTS_tt_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id"))  %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

@@

tld_2011_nonmotor <- od_trips_2011_nonmotor |> filter(travel_time <= 27) |>
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2016_nonmotor <- od_trips_2016_nonmotor |> filter(travel_time <= 27) |>
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))
```

@@

all_tt_2011_nonmotor <- od_trips_2011_nonmotor |> filter(travel_time <= 27) |>
  dplyr::select(Trips, travel_time)
sum((all_tt_2011_nonmotor$Trips))
all_tt_2011_nonmotor <- all_tt_2011_nonmotor[rep(seq_len(dim(all_tt_2011_nonmotor)[1]), all_tt_2011_nonmotor$Trips), 2]

@@

all_tt_2016_nonmotor <- od_trips_2016_nonmotor  |> filter(travel_time <= 27) |>
  dplyr::select(Trips, travel_time)
sum((all_tt_2016_nonmotor$Trips))
all_tt_2016_nonmotor <- all_tt_2016_nonmotor[rep(seq_len(dim(all_tt_2016_nonmotor)[1]), all_tt_2016_nonmotor$Trips), 2]

@@

And save the gamma functions, to use in the analysis
```{r}
save(gamma2011_motor, file="data-products/TTS_gamma2011_motor.Rdata")
save(gamma2016_motor, file="data-products/TTS_gamma2016_motor.Rdata")
save(exp2011_nonmotor, file="data-products/TTS_exp2011_nonmotor.Rdata")
save(exp2016_nonmotor, file="data-products/TTS_exp2016_nonmotor.Rdata")
```
